---
title: "Time random effects in CJS models: TMB for the win"
output: html_document
date: "2023-08-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Motivation

I'd like to have temporal random effect on the recapture probability of the CJS model to account for variation in sampling effort. I could go for a temporal covariate, a proxy of sampling effort. But a time effect should do the job too. A fixed time effect is too data hungry. Let's explore random effects. The reflex is to go Bayesian, with NIMBLE as far as I'm concerned, but I'd like to explore a frequentist approach as well, with TMB.

## Nice resources to introduce TMB with R

+ A very nice intro by Rebecca Stubbs, perfect for beginners: https://rpubs.com/BeccaStubbs/tmb_simple_distributions

+ Another very nice intro by the Eco-Stats research group https://www.dropbox.com/s/f5ng7t6pgnvcvmp/TMB_Intro.html?dl=0

+ In the HMM context, there is this cool paper by Bacri and colleagues: https://onlinelibrary.wiley.com/doi/10.1002/bimj.202100256

+ Some official useful material:   
    - THe TMB wiki: https://github.com/kaskr/adcomp/wiki
    - The TMB book: https://kaskr.github.io/adcomp/_book/Introduction.html
    - The official introduction/tutorial to TMB https://github.com/kaskr/adcomp/wiki/Tutorial
    + The material from a TMB workshop: https://github.com/TMB-ADMB-Workshops/tmb_workshop

## Simple model with constant survival/recapture probabilities

We use the dipper data for illustration. 

Wrangle data:
```{r}
# get dipper data
library(tidyverse)
library(mra)
data(dipper.data)
data <- dipper.data %>% select(-males,-females) %>% as.matrix()

# define various quantities
nh <- dim(data)[1]
k <- dim(data)[2]
km1 <- k-1

# counts
eff <- rep(1,nh)
  
# compute the date of first capture fc, and state at initial capture init.state
fc <- NULL
init.state <- NULL
for (i in 1:nh){
  temp <- 1:k
  fc <- c(fc,min(which(data[i,]!=0)))
  init.state <- c(init.state,data[i,fc[i]])
}

# transpose data
data <- t(data)
```

Now the TMB implementation:
```{r}
# load TMB
library(TMB)

# delete existing files before compilation
cpp_version <- "phip2"
if (file.exists(paste0(cpp_version,".o"))) file.remove(paste0(cpp_version,".so"),
                                                       paste0(cpp_version,".o"))

# compile and load library
compile("phip2.cpp")
dyn.load(dynlib("phip2"))
```

Optimisation:
```{r message=FALSE, warning=FALSE}
f <- MakeADFun(
  data = list(ch = data, fc = fc, fs = init.state), 
  parameters = list(phib = 0.5, detb = 0.5),
  DLL = "phip2",
  silent = TRUE)
opt <- do.call("optim", f) # optimisation
opt$par
rep <- sdreport(f)
plogis(rep$par.fixed)
```

## Model with fixed effect of time on recapture prob

Get/wrangle data:
```{r}
# get dipper data
library(tidyverse)
library(mra)
data(dipper.data)
data <- dipper.data %>% select(-males,-females) %>% as.matrix()

# define various quantities
nh <- dim(data)[1]
k <- dim(data)[2]
km1 <- k-1

# counts
eff <- rep(1,nh)
  
# compute the date of first capture fc, and state at initial capture init.state
fc <- NULL
init.state <- NULL
for (i in 1:nh){
  temp <- 1:k
  fc <- c(fc,min(which(data[i,]!=0)))
  init.state <- c(init.state,data[i,fc[i]])
}

# transpose data
data <- t(data)
```

Now the TMB implementation:
```{r}
library(TMB)
cpp_version <- "phipt"
if (file.exists(paste0(cpp_version,".o"))) file.remove(paste0(cpp_version,".so"),
                                                       paste0(cpp_version,".o"))
compile("phipt.cpp")
dyn.load(dynlib("phipt"))
```

Optimisation:
```{r message=FALSE, warning=FALSE}
parameters <- list(phib = runif(1), detb = runif(6))
mydata <- list(ch = data, fc = fc, fs = init.state)
obj <- MakeADFun(data = mydata, parameters = parameters, DLL = "phipt")
obj$hessian <- TRUE
opt <- do.call("optim", obj)
opt
opt$hessian ## <-- FD hessian from optim
obj$he()    ## <-- Analytical hessian

rep <- sdreport(obj)
rep
plogis(rep$par.fixed)
summary(rep, "report")
```


With RMark:
```{r}
library(RMark)
dipper <- data.frame(ch = unite(t(data) %>% as_tibble(), 
                                col = "ch", 
                                sep = ""))
dipper.proc <- process.data(data = dipper)
dipper.ddl <- make.design.data(dipper.proc)
phi <- list(formula=~1)     # constant
pt <- list(formula=~time)   # time
phi.pt <- mark(dipper.proc,
               dipper.ddl,
               model.parameters = list(Phi = phi, p = pt),
               delete = TRUE)
phi.pt$results$real

```

We're good.


## Model with random effect of time on recapture prob

### Simulations

I simulate encounter histories starting with 650 individuals over 20 years with survival set to 0.8. I release 200 individuals every 5 years, except the last cohort which is 250 individuals. Detection is 0.6 the first 15 years, then 0.9 for the 4 remaining years:
```{r}
# set the scene
set.seed(3) # for reproducibility
nyear <- 20 # duration of the study
det <- c(rep(0.6,15), rep(0.9,4))# detection
phi <- 0.8
nind <- 650 # nb of individuals
expit <- function(x){exp(x)/(1+exp(x))} # reciprocal logit function
z <- data <- x <- matrix(NA, nrow = nind, ncol = nyear)
first <- c(rep(1, 200), rep(5, 200), rep(10, 200), rep(15, 250))

# simulate encounter histories
for(i in 1:nind){
  z[i,first[i]] <- x[i,first[i]] <- 1
  for(j in (first[i]+1):nyear){
    z[i,j]<-rbinom(1,1,phi*z[i,j-1])
    x[i,j]<-rbinom(1,1,z[i,j]*det[j-1])
  }
}
his <- x
his[is.na(his)] <- 0 # remove lines with 0's
```


### RMARK

```{r}
# format data
k <- ncol(his) # nb of capture occasions
n <- nrow(his) # nb of individuals
out <- array(dim=n)
for (i in 1:n){
	y <- (his[i,] > 0) * 1
	out[i] <- paste(y,collapse="")
}
capt.hist <- data.frame(ch = out)

# model fitting
library(RMark)
dipper.proc <- process.data(data = capt.hist)
dipper.ddl <- make.design.data(dipper.proc)
phi <- list(formula=~1)     # constant
pt <- list(formula=~time)   # time
phi.pt <- mark(dipper.proc,
               dipper.ddl,
               model.parameters = list(Phi = phi, p = pt),
               delete = TRUE)
#phi.pt$results$beta
phi.pt$results$real
```

### NIMBLE

```{r}
# load NIMBLE
library(nimble)

# Model code
hmm.phipt <- nimbleCode({
  phi ~ dunif(0, 1) # prior survival
  gamma[1,1] <- phi      # Pr(alive t -> alive t+1)
  gamma[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
  gamma[2,1] <- 0        # Pr(dead t -> alive t+1)
  gamma[2,2] <- 1        # Pr(dead t -> dead t+1)
  delta[1] <- 1          # Pr(alive t = 1) = 1
  delta[2] <- 0          # Pr(dead t = 1) = 0
  sig ~ dunif(0, 5)
  mu ~ dnorm(0, 1.5)
  for (t in 1:(T-1)){
    logit(p[t]) <- mu + sig * eps[t] # prior detection
    eps[t] ~ dnorm(0, sd = 1)
    omega[1,1,t] <- 1 - p[t]    # Pr(alive t -> non-detected t)
    omega[1,2,t] <- p[t]        # Pr(alive t -> detected t)
    omega[2,1,t] <- 1        # Pr(dead t -> non-detected t)
    omega[2,2,t] <- 0        # Pr(dead t -> detected t)
  }
  # likelihood
  for (i in 1:N){
    z[i,first[i]] ~ dcat(delta[1:2])
    for (j in (first[i]+1):T){
      z[i,j] ~ dcat(gamma[z[i,j-1], 1:2])
      y[i,j] ~ dcat(omega[z[i,j], 1:2, j-1])
    }
  }
})
y <- his
# occasions of first capture
first <- apply(y, 1, function(x) min(which(x !=0)))
# constants
my.constants <- list(N = nrow(y), 
                     T = ncol(y), 
                     first = first)
# data
my.data <- list(y = y + 1)
# initial values
zinits <- y + 1 # non-detection -> alive
zinits[zinits == 2] <- 1 # dead -> alive
initial.values <- function() list(phi = runif(1,0,1),
                                  mu = rnorm(1, 0, 1),
                                  sig = runif(1,0,5),
                                  z = zinits)
# parameters to monitor
parameters.to.save <- c("phi", "mu", "sig", "p")
# MCMC details
n.iter <- 10000
n.burnin <- 5000
n.chains <- 2
# run NIMBLE
mcmc.phipt <- nimbleMCMC(code = hmm.phipt, 
                        constants = my.constants,
                        data = my.data,              
                        inits = initial.values,
                        monitors = parameters.to.save,
                        niter = n.iter,
                        nburnin = n.burnin, 
                        nchains = n.chains)
# numerical summaries
MCMCvis::MCMCsummary(mcmc.phipt, round = 2)
```

### TMB

```{r}
# get simulated data
data <- his

# define various quantities
nh <- dim(data)[1]
k <- dim(data)[2]
km1 <- k-1

# counts
eff <- rep(1,nh)

# compute the date of first capture fc, and state at initial capture init.state
fc <- NULL
init.state <- NULL
for (i in 1:nh){
  temp <- 1:k
  fc <- c(fc,min(which(data[i,]!=0)))
  init.state <- c(init.state,data[i,fc[i]])
}

# transpose data
data <- t(data)

# load TMB
library(TMB)

# delete existing files before compilation
cpp_version <- "phipREt"
if (file.exists(paste0(cpp_version,".o"))) file.remove(paste0(cpp_version,".so"),
                                                       paste0(cpp_version,".o"))

# compile and load library
compile("phipREt.cpp")
dyn.load(dynlib("phipREt"))

# inits
parameters <- list(phib = 0.8,
                   mu = 0,
                   logsig = 0,
                   u = rep(0, 19))
# data
mydata <- list(ch = data, fc = fc, fs = init.state)

# optimisation
obj <- MakeADFun(data = mydata, 
                 parameters = parameters,
                 random = "u",
                 DLL = "phipREt")
obj$hessian <- TRUE
opt <- optim(par = obj$par, fn = obj$fn, gr = obj$gr, method = "BFGS", 
             control = list(maxit = 1000), hessian = T)

rep <- sdreport(obj)
rep

summary(rep, "report")
```

